---
title: "Individual Differences in Degree of Discounting"
subtitle: "Computational Appendix for Wan, Myerson, & Green (2023)"
author: "Haoran (Matt) Wan"
date: "today"
format: 
  html:
    toc: true
    code-fold: true
    self-contained: true
engine: knitr
---

## Introduction

This document contains the R code to replicate the primary analyses from the publication:

> Wan, H., Myerson, J., & Green, L. (2023). Individual differences in degree of discounting: Do different procedures and measures assess the same construct?. *Behavioural Processes*, *208*, 104864. https://doi.org/10.1016/j.beproc.2023.104864

The analyses compare two methods for measuring delay discounting—the Adjusting-Amount (Adj-Amt) procedure and the Monetary Choice Questionnaire (MCQ) using a variety of statistical techniques, including nonlinear regression, beta regression, correlation, and ANOVA.

The data can be requested from me or from my coauthors, Leonard Green and Joel Myerson.

## 1. Setup: Load Packages and Process Data

This chunk loads all necessary R packages and then performs the data processing steps, including calculating the theoretical (`log k`) and atheoretical (Area under the Curve / choice proportion) discounting measures for each individual.

```{r setup_and_process}
#| message: false
#| warning: false

# --- Load Packages ---
library(minpack.lm)
library(glmmTMB)
library(multcomp)
library(emmeans)
library(psych)
library(readr)
library(dplyr)
library(tidyr)

# --- Load and Process Data (from Data.R) ---
disc_df <- read_csv("R Code/data/AdjAmt_MCQ.csv") |> select(-1)

# --- Group-Level Data Frame for Plotting/Group Fits ---
disc_grp_df <- disc_df |>
  group_by(provider, procedure, amt, iv) |>
  mutate(iv = ifelse(procedure == "mcq", log(iv), iv)) |>
  summarise(mean_sv = mean(value), med_sv = median(value), .groups = 'drop') |>
  arrange(procedure, provider)

# --- Individual-Level Data Frame ---
# Function to calculate Area under the Curve
auc <- function(x, y) {
  sum(diff(x) * (y[-1] + y[-length(y)])) / 2
}

# Process Adj-Amt Data
adj_amt_processed <- disc_df |>
  filter(procedure == "aa" & iv != 730) |>
  group_by(id, procedure, amt) |>
  summarise(
    atheoretical = auc(iv / 180, value),
    theoretical = tryCatch({
      coef(nlsLM(log(value) ~ -log(1 + exp(k) * iv), start = list(k = -4), 
                  control = nls.lm.control(maxiter = 1024)))[['k']]
    }, error = function(e) NA),
    .groups = 'drop'
  ) |>
  select(id, procedure, amt, atheoretical, theoretical)

# Process MCQ Data
# Using the original, more robust logic from Data.R to calculate theoretical log k
mcq_processed <- disc_df |>
  filter(procedure == "mcq") |>
  group_by(id, amt) |>
  summarise(
    # atheoretical measure is the sum of choices (0 or 1), which is correct.
    atheoretical = sum(value),
    # theoretical measure requires finding the indifference point (crossover)
    theoretical = {
      # Local variables for clarity within the summarise call
      choices <- value
      k_values <- iv
      # Handle edge cases first for participants who never switch 
      if (all(choices == 1)) {
        # If all choices are for the delayed option, use the smallest k
        log(0.00016)
      } else if (all(choices == 0)) {
        # If all choices are for the immediate option, use the largest k
        log(0.25)
      } else {
        # For participants who switch, calculate consistency for each k
        n_consistent <- sapply(k_values, function(k_level) {
          sum((choices == 0 & k_values <= k_level) | (choices == 1 & k_values >= k_level))
        })
        # Find ALL k-values that maximize the number of consistent choices
        indifference_ks <- k_values[which(n_consistent == max(n_consistent))]
        # The estimated indifference point is the geometric mean of these k's
        log(geometric.mean(indifference_ks))
      }
    },
    .groups = 'drop'
  ) |>
  # Add the procedure column back for merging
  mutate(procedure = "mcq") |>
  # Ensure only one row per participant/amount combination
  distinct(id, amt, .keep_all = TRUE)

# Combine into final analysis dataframe
behav <- bind_rows(adj_amt_processed, mcq_processed) |>
    left_join(distinct(select(disc_df, id, provider)), by = "id")
```

> **Note:** The `log k` values (the `theoretical` measure) in this script are calculated using a direct scoring algorithm (for the MCQ) and nonlinear least-squares (`nlsLM` in R) for consistency within this computational environment. The values reported in the published article were estimated using Bayesian methods in Stan, which may result in minor numerical differences.

---

## 2. Results

The following analyses are structured to match the Results section of the published article.

### 2.1 Group-Level Analyses & Amount Effects

This section replicates the group-level model fits and tests for the magnitude effect, where the degree of delay discounting decreases as the reward amount increases.

```{r group_analysis}
#| warning: false

# --- Adjusting-Amount Procedure ---
cat("--- Adj-Amt: Group-level Hyperboloid Model Fit (R-squared) ---\n")
r2_aa_grp <- matrix(NA, nrow = 3, ncol = 2) |>
  `colnames<-`(c("Prolific", "MTurk")) |>
  `rownames<-`(c("$30", "$80", "$500"))

for (a in 1:3) {
  for (p in c("Prolific", "MTurk")) {
    data_subset <- subset(disc_grp_df, amt == a & procedure == "aa" & provider == p)
    fit <- nlsLM(med_sv ~ 1 / (1 + exp(k) * iv)^b, data = data_subset, 
                 start = list(k = -4, b = 1), control = list(maxiter = 1000))
    rss <- sum(residuals(fit)^2)
    tss <- sum((data_subset$med_sv - mean(data_subset$med_sv))^2)
    r2_aa_grp[a, p] <- 1 - (rss / tss)
  }
}
print(r2_aa_grp, digits = 3)

cat("\n--- Adj-Amt: Amount Effect (Beta Regression p-values) ---\n")
behav |> filter(procedure == "aa") |>
  group_by(provider) |>
  summarise(
    p_value = summary(glht(
      glmmTMB(atheoretical ~ as.factor(amt) + (1 | id), family = beta_family(), data = cur_data()),
      linfct = matrix(c(-1, 0, 1), nrow = 1)
    ))$test$pvalues[1],
    .groups = 'drop'
  )

# --- MCQ Procedure ---
cat("\n--- MCQ: Group-level Logistic Growth Model Fit (R-squared) ---\n")
r2_mcq_grp <- matrix(NA, nrow = 3, ncol = 2) |>
  `colnames<-`(c("Prolific", "MTurk")) |>
  `rownames<-`(c("$30", "$55", "$80"))

for (a in 1:3) {
  for (p in c("Prolific", "MTurk")) {
    data_subset <- subset(disc_grp_df, amt == a & procedure == "mcq" & provider == p)
    fit <- nlsLM(mean_sv ~ 1 / (1 + exp(-(iv - x) * r)), data = data_subset, 
                 start = list(x = -4, r = 1), control = list(maxiter = 1000))
    rss <- sum(residuals(fit)^2)
    tss <- sum((data_subset$mean_sv - mean(data_subset$mean_sv))^2)
    r2_mcq_grp[a, p] <- 1 - (rss / tss)
  }
}
print(r2_mcq_grp, digits = 3)

cat("\n--- MCQ: Amount Effect (Logistic Regression p-values) ---\n")
behav |> filter(procedure == "mcq") |>
  group_by(provider) |>
  summarise(
    p_value = summary(glht(
      glmmTMB(cbind(atheoretical, 9 - atheoretical) ~ as.factor(amt) + (1 | id), family = binomial, data = cur_data()),
      linfct = matrix(c(contr.poly(3)[,1]), nrow = 1)
    ))$test$pvalues[1],
    .groups = 'drop'
  )
```

---

### 2.2 Reliability: Within-Procedure Correlations

These analyses evaluate the reliability of the discounting measures by examining the correlations between different reward amounts within each procedure. The results correspond to **Tables 1 and 2** in the paper.

```{r reliability_analysis}
# --- Correlation within each discounting measure ---
cat("--- Within-Procedure, Within-Measure Correlations ---\n")
behav |>
  pivot_longer(names_to = "measure", values_to = "value", cols = c(atheoretical, theoretical)) |>
  pivot_wider(names_from = amt, values_from = value) |>
  group_by(measure, procedure, provider) |>
  summarise(
    cor_1_2 = cor(`1`, `2`, use = "complete.obs"),
    cor_1_3 = cor(`1`, `3`, use = "complete.obs"),
    cor_2_3 = cor(`2`, `3`, use = "complete.obs"),
    .groups = 'drop'
  ) |> print(n=20)

# --- Correlation between discounting measures (theoretical and atheoretical) ---
cat("\n--- Within-Procedure, Between-Measure Correlations ---\n")
behav |>
  group_by(provider, procedure, amt) |>
  summarise(correlation = cor(atheoretical, theoretical, use = "complete.obs"), .groups = 'drop') |>
  pivot_wider(names_from = amt, values_from = correlation, names_prefix = "amt_")
```

---

### 2.3 Construct Validity: Between-Procedure Correlations

This analysis addresses the primary research question by correlating the measures from the Adj-Amt procedure with those from the MCQ. This corresponds to **Table 3** in the paper.

```{r validity_analysis}
cat("--- Between-Procedure Correlations (Construct Validity) ---\n")
behav |>
  filter((procedure == "aa" & amt != 3) | (procedure == "mcq" & amt != 2)) |>
  mutate(amt = ifelse(amt == 1, "$30", "$80")) |>
  pivot_wider(names_from = procedure, values_from = c(atheoretical, theoretical)) |>
  group_by(provider, amt) |> 
  summarise(
    cor_atheoretical = cor(atheoretical_aa, atheoretical_mcq, use = "complete.obs"),
    cor_theoretical = cor(theoretical_aa, theoretical_mcq, use = "complete.obs"),
    .groups = 'drop'
  )
```

---

### 2.4 Comparing Degrees of Discounting

An ANOVA was conducted to compare the absolute `log k` values between procedures, samples (Prolific vs. MTurk), and reward amounts.

```{r anova_analysis}
cat("--- ANOVA: Comparing log k Across Procedures, Samples, and Amounts ---\n")
# Prepare data for ANOVA
anova_data <- filter(behav, (procedure == "aa" & amt != 3) | (procedure == "mcq" & amt != 2)) |>
  mutate(
    amt = factor(ifelse(amt == 1, "$30", "$80")),
    procedure = factor(procedure),
    provider = factor(provider)
  )

# Fit and print ANOVA summary
logk_mod <- aov(theoretical ~ provider * amt * procedure, data = anova_data)
print(summary(logk_mod))

cat("\n--- Post-Hoc Contrasts for Procedure x Amount Interaction ---\n")
emm_obj <- emmeans(logk_mod, ~ amt | procedure)
print(pairs(emm_obj, adjust = "holm"))
```