---
title: "Do the Adjusting-Amount Procedure and MCQ Measure the Same Construct?"
subtitle: "R reproducibility companion to Wan, Myerson, and Green (2023)"
author: "Haoran Wan"
date: "2026-07-01"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    self-contained: true
    theme: cosmo
execute:
  warning: false
  message: false
engine: knitr
---

## Executive summary

This report examines whether two widely used measures of delay discounting - the **Adjusting-Amount procedure (Adj-Amt)** and the **Monetary Choice Questionnaire (MCQ)** - capture the same individual-difference construct.

The published study analyzed 393 adults recruited through Prolific (*n* = 150) and Amazon Mechanical Turk (MTurk; *n* = 243 after attention-check exclusions). Three findings are central:

1. **Both procedures produced the expected magnitude effect.** Larger delayed rewards were discounted less steeply in both samples and with both procedures (all published *p*s < .001).
2. **Individual differences were reliable and convergent.** Published between-procedure correlations ranged from .741 to .821 across samples, reward amounts, and scoring approaches.
3. **Convergence does not imply interchangeability.** At $30, the MCQ produced steeper discounting than Adj-Amt (*p* = .003); at $80, the procedures did not differ (*p* = .95).

::: {.callout-important}
## Scientific conclusion

Adj-Amt and MCQ scores capture substantial common variance and are consistent with a shared underlying construct. Their `log k` estimates should not, however, be treated as numerically interchangeable without considering reward amount, procedure, and recruitment source.
:::

## Published study

> Wan, H., Myerson, J., & Green, L. (2023). Individual differences in degree of discounting: Do different procedures and measures assess the same construct? *Behavioural Processes, 208*, 104864. <https://doi.org/10.1016/j.beproc.2023.104864>

### Research questions

The article addressed four questions in sequence:

1. Do both procedures reproduce the magnitude effect?
2. Are individual differences reliable across reward amounts and scoring approaches?
3. Do Adj-Amt and MCQ measure the same construct?
4. Do the procedures and online samples yield the same absolute degree of discounting?

### Design and sample

| Feature | Prolific | MTurk |
|---|---:|---:|
| Recruited | 150 | 275 |
| Excluded for failing at least one of seven attention checks | 0 | 32 |
| Analyzed | 150 | 243 |
| Mean age (SD) | 28.5 (8.5) | 34.8 (10.4) |
| Age range | 19-72 | 20-70 |
| Mean completion time | 10.1 min | 10.2 min |

Participants completed both procedures in counterbalanced order. The Adj-Amt procedure measured subjective values for delayed rewards of $30, $80, and $500. The MCQ contained three nine-item sets with delayed amounts of $25-$35, $50-$60, and $75-$85.

### Measures

| Procedure | Atheoretical measure | Theoretically based measure |
|---|---|---|
| Adj-Amt | Area under the empirical discounting curve (AuC) | `log k` from a simple hyperbola |
| MCQ | Proportion of larger-delayed choices | `log k` from Kirby's consistency-scoring method |

Lower AuC and delayed-choice proportions indicate steeper discounting. Higher `log k` indicates steeper discounting; correlations between atheoretical measures and `log k` are therefore expected to be negative.

::: {.callout-note}
## Reproducibility boundary

The archived project code indicates that the publication's Adj-Amt `log k` estimates were obtained with a Stan model that treated preference reversals probabilistically; the article describes these estimates as simple-hyperbola fits. This companion report estimates Adj-Amt `log k` by nonlinear least squares from the analysis-ready subjective values. The code therefore reproduces the analysis logic and substantive pattern, but not every published theoretical point estimate. Published statistics are shown explicitly as benchmarks and are never relabeled as newly computed results.
:::

::: {.callout-note}
## Data availability

Participant-level data are not included in this repository. The analysis-ready file is available from the study authors on request. Authorized recipients should save it as `Analysis/AdjAmt_MCQ.csv` before executing this report.
:::

## 1. Setup and data integrity

```{r setup}
library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(minpack.lm)
library(glmmTMB)
library(emmeans)
library(knitr)

options(digits = 3, scipen = 4)

if (!file.exists("AdjAmt_MCQ.csv")) {
  stop(
    paste(
      "Participant-level data are not included in this repository.",
      "Request the analysis-ready file from the study authors and save it as",
      "Analysis/AdjAmt_MCQ.csv before running this report."
    ),
    call. = FALSE
  )
}

raw_data <- read_csv("AdjAmt_MCQ.csv", show_col_types = FALSE) |>
  select(-1)

stopifnot(
  nrow(raw_data) == 15720,
  n_distinct(raw_data$id) == 393,
  setequal(unique(raw_data$provider), c("Prolific", "MTurk")),
  setequal(unique(raw_data$procedure), c("aa", "mcq"))
)

participant_data <- raw_data |>
  distinct(id, provider, age, sex, edu, eth, race)

sample_summary <- participant_data |>
  group_by(provider) |>
  summarise(
    n = n(),
    age_mean = mean(age, na.rm = T),
    age_sd = sd(age, na.rm = T),
    age_range = paste0(min(age, na.rm = T), "-", max(age, na.rm = T)),
    male_percent = 100 * mean(sex == 0, na.rm = TRUE),
    below_college_percent = 100 * mean(edu == 0, na.rm = TRUE),
    hispanic_latino_percent = 100 * mean(eth == 0, na.rm = TRUE),
    non_white_percent = 100 * mean(race == 0, na.rm = TRUE),
    .groups = "drop"
  )

kable(sample_summary, digits = 1, caption = "Sample characteristics in the analysis-ready dataset.")
```

When supplied by the authors, the analysis-ready file contains only participants retained for analysis. Attention-check failures must therefore be documented from the published recruitment flow rather than recomputed from this file.

### Scoring functions

```{r scoring-functions}
calculate_auc <- function(delay, subjective_value) {
  order_index <- order(delay)
  x <- delay[order_index] / 180
  y <- subjective_value[order_index]
  sum(diff(x) * (y[-1] + y[-length(y)]) / 2)
}

fit_adj_logk <- function(delay, subjective_value) {
  fit <- tryCatch(
    nlsLM(
      log(subjective_value) ~ -log(1 + exp(log_k) * delay),
      start = list(log_k = -4),
      control = nls.lm.control(maxiter = 1024)
    ),
    error = function(e) NULL
  )
  if (is.null(fit)) NA_real_ else unname(coef(fit)[["log_k"]])
}

score_mcq_logk <- function(question_k, delayed_choice) {
  consistency <- vapply(
    question_k,
    function(candidate_k) {
      sum(
        (delayed_choice == 0 & question_k <= candidate_k) |
          (delayed_choice == 1 & question_k >= candidate_k)
      )
    },
    numeric(1)
  )
  best_k <- question_k[consistency == max(consistency)]
  log(exp(mean(log(best_k))))
}
```

```{r participant-scores}
adj_scores <- raw_data |>
  filter(procedure == "aa", iv != 730) |>
  arrange(id, amt, iv) |>
  group_by(id, provider, amt) |>
  summarise(
    auc = calculate_auc(iv, value),
    log_k_companion = fit_adj_logk(iv, value),
    .groups = "drop"
  )

mcq_scores <- raw_data |>
  filter(procedure == "mcq") |>
  arrange(id, amt, iv) |>
  group_by(id, provider, amt) |>
  summarise(
    delayed_choices = sum(value),
    delayed_choice_prop = mean(value),
    log_k_companion = score_mcq_logk(iv, value),
    .groups = "drop"
  )

stopifnot(
  nrow(adj_scores) == 393 * 3,
  nrow(mcq_scores) == 393 * 3,
  all(mcq_scores$delayed_choices %in% 0:9),
  all(mcq_scores$delayed_choice_prop >= 0 & mcq_scores$delayed_choice_prop <= 1)
)
```

## 2. Benchmark validity: the magnitude effect

Larger delayed rewards are typically discounted less steeply than smaller delayed rewards. Reproducing this magnitude effect establishes that both procedures generated the expected behavioral pattern.

### Group-level model fit

```{r group-fits}
group_summary <- raw_data |>
  group_by(provider, procedure, amt, iv) |>
  summarise(
    mean_value = mean(value),
    median_value = median(value),
    .groups = "drop"
  )

adj_group_fits <- group_summary |>
  filter(procedure == "aa") |>
  group_by(provider, amt) |>
  group_modify(~ {
    fit <- nlsLM(
      median_value ~ 1 / (1 + exp(log_k) * iv)^s,
      data = .x,
      start = list(log_k = -4, s = 1),
      control = nls.lm.control(maxiter = 1024)
    )
    tibble(r_squared = 1 - sum(residuals(fit)^2) /
      sum((.x$median_value - mean(.x$median_value))^2))
  }) |>
  mutate(amount = recode(as.character(amt), `1` = "$30", `2` = "$80", `3` = "$500")) |>
  select(provider, amount, r_squared)

mcq_group_fits <- group_summary |>
  filter(procedure == "mcq") |>
  mutate(log_question_k = log(iv)) |>
  group_by(provider, amt) |>
  group_modify(~ {
    fit <- nlsLM(
      mean_value ~ 1 / (1 + exp(-(log_question_k - intercept) * rate)),
      data = .x,
      start = list(intercept = -4, rate = 1),
      control = nls.lm.control(maxiter = 1024)
    )
    tibble(r_squared = 1 - sum(residuals(fit)^2) /
      sum((.x$mean_value - mean(.x$mean_value))^2))
  }) |>
  mutate(amount = recode(
    as.character(amt),
    `1` = "$25-$35", `2` = "$50-$60", `3` = "$75-$85"
  )) |>
  select(provider, amount, r_squared)

kable(adj_group_fits, digits = 3, caption = "Companion group-level hyperboloid fits for Adj-Amt.")
kable(mcq_group_fits, digits = 3, caption = "Companion group-level logistic fits for MCQ.")
```

### Magnitude-effect tests

The published analyses used beta regression for Adj-Amt AuC and logistic regression for MCQ delayed-choice counts, with linear contrasts across amount. The following models preserve that outcome-specific approach and account for repeated observations with participant random intercepts.

```{r amount-effects}
adj_amount_effects <- adj_scores |>
  mutate(amount = factor(amt, levels = 1:3)) |>
  split(.$provider) |>
  imap_dfr(function(data, sample_name) {
    model <- glmmTMB(
      auc ~ amount + (1 | id),
      family = beta_family(),
      data = data
    )
    test <- contrast(emmeans(model, ~ amount), method = "poly") |>
      as.data.frame()
    tibble(
      provider = sample_name,
      linear_trend = test$estimate[1],
      p_value = test$p.value[1]
    )
  })

mcq_amount_effects <- mcq_scores |>
  mutate(amount = factor(amt, levels = 1:3)) |>
  split(.$provider) |>
  imap_dfr(function(data, sample_name) {
    model <- glmmTMB(
      cbind(delayed_choices, 9 - delayed_choices) ~ amount + (1 | id),
      family = binomial(),
      data = data
    )
    test <- contrast(emmeans(model, ~ amount), method = "poly") |>
      as.data.frame()
    tibble(
      provider = sample_name,
      linear_trend = test$estimate[1],
      p_value = test$p.value[1]
    )
  })

kable(adj_amount_effects, digits = 4, caption = "Adj-Amt magnitude-effect tests.")
kable(mcq_amount_effects, digits = 4, caption = "MCQ magnitude-effect tests.")
```

The article reported a significant magnitude effect in every procedure-by-sample analysis (all *p*s < .001).

## 3. Reliability within each procedure

The article treated scores at different reward amounts as alternative forms. Large positive correlations across amounts therefore indicate that the procedures preserve individual differences across changes in reward magnitude.

```{r reliability}
amount_correlations <- function(data, measure) {
  data |>
    select(id, provider, amt, {{ measure }}) |>
    pivot_wider(names_from = amt, values_from = {{ measure }}, names_prefix = "amount_") |>
    group_by(provider) |>
    summarise(
      `small-medium` = cor(amount_1, amount_2, use = "complete.obs"),
      `small-large` = cor(amount_1, amount_3, use = "complete.obs"),
      `medium-large` = cor(amount_2, amount_3, use = "complete.obs"),
      .groups = "drop"
    )
}

adj_auc_reliability <- amount_correlations(adj_scores, auc) |>
  mutate(procedure = "Adj-Amt", measure = "AuC")
adj_logk_reliability <- amount_correlations(adj_scores, log_k_companion) |>
  mutate(procedure = "Adj-Amt", measure = "Companion log k")
mcq_prop_reliability <- amount_correlations(mcq_scores, delayed_choice_prop) |>
  mutate(procedure = "MCQ", measure = "Delayed-choice proportion")
mcq_logk_reliability <- amount_correlations(mcq_scores, log_k_companion) |>
  mutate(procedure = "MCQ", measure = "Companion log k")

companion_reliability <- bind_rows(
  adj_auc_reliability,
  adj_logk_reliability,
  mcq_prop_reliability,
  mcq_logk_reliability
) |>
  select(procedure, measure, provider, everything())

kable(companion_reliability, digits = 3, caption = "Companion alternate-forms correlations.")
```

```{r measure-convergence}
measure_convergence <- bind_rows(
  adj_scores |>
    group_by(provider, amt) |>
    summarise(r = cor(auc, log_k_companion), .groups = "drop") |>
    mutate(procedure = "Adj-Amt"),
  mcq_scores |>
    group_by(provider, amt) |>
    summarise(r = cor(delayed_choice_prop, log_k_companion), .groups = "drop") |>
    mutate(procedure = "MCQ")
) |>
  mutate(amount = case_when(
    procedure == "Adj-Amt" & amt == 1 ~ "$30",
    procedure == "Adj-Amt" & amt == 2 ~ "$80",
    procedure == "Adj-Amt" & amt == 3 ~ "$500",
    procedure == "MCQ" & amt == 1 ~ "$25-$35",
    procedure == "MCQ" & amt == 2 ~ "$50-$60",
    TRUE ~ "$75-$85"
  )) |>
  select(procedure, provider, amount, r)

kable(measure_convergence, digits = 3,
  caption = "Correlations between atheoretical measures and companion log k estimates.")
```

The published within-procedure correlations were uniformly significant and ranged from .737 to .925 across amounts. Correlations between atheoretical and theoretical measures were strongly negative, as expected from their opposite scoring directions.

## 4. Convergent validity between procedures

Only two amount conditions were comparable across procedures: Adj-Amt $30 with MCQ $25-$35, and Adj-Amt $80 with MCQ $75-$85.

### Published between-procedure correlations (Table 3)

| Sample | Common amount | `log k` | AuC with delayed-choice proportion |
|---|---:|---:|---:|
| Prolific | $30 | .768 | .809 |
| Prolific | $80 | .788 | .821 |
| MTurk | $30 | .741 | .790 |
| MTurk | $80 | .770 | .820 |

```{r between-procedure}
adj_common <- adj_scores |>
  filter(amt %in% c(1, 2)) |>
  mutate(amount = if_else(amt == 1, "$30", "$80")) |>
  transmute(id, provider, amount, auc, adj_log_k = log_k_companion)

mcq_common <- mcq_scores |>
  filter(amt %in% c(1, 3)) |>
  mutate(amount = if_else(amt == 1, "$30", "$80")) |>
  transmute(
    id, provider, amount,
    delayed_choice_prop,
    mcq_log_k = log_k_companion
  )

common_scores <- inner_join(
  adj_common,
  mcq_common,
  by = c("id", "provider", "amount")
)

companion_between <- common_scores |>
  group_by(provider, amount) |>
  summarise(
    companion_atheoretical_r = cor(auc, delayed_choice_prop),
    companion_log_k_r = cor(adj_log_k, mcq_log_k),
    .groups = "drop"
  )

published_between <- tribble(
  ~provider, ~amount, ~published_atheoretical_r, ~published_log_k_r,
  "Prolific", "$30", .809, .768,
  "Prolific", "$80", .821, .788,
  "MTurk", "$30", .790, .741,
  "MTurk", "$80", .820, .770
)

between_comparison <- left_join(
  published_between,
  companion_between,
  by = c("provider", "amount")
)

kable(between_comparison, digits = 3,
  caption = "Published benchmarks and companion between-procedure correlations.")
```

The atheoretical results reproduce the published values because they are calculated directly from the same observed responses. Small differences in the theoretical correlations are expected because the companion Adj-Amt estimates use nonlinear least squares rather than the publication's Stan model.

## 5. Absolute differences: convergence is not equivalence

The publication compared `log k` values with a 2 (sample) x 2 (amount) x 2 (procedure) factorial ANOVA. The companion model below uses the transparent nonlinear-least-squares estimates; the published ANOVA remains the inferential reference.

```{r factorial-anova}
companion_anova_data <- bind_rows(
  adj_scores |>
    filter(amt %in% c(1, 2)) |>
    transmute(
      id, provider,
      amount = factor(if_else(amt == 1, "$30", "$80")),
      procedure = factor("Adj-Amt"),
      log_k = log_k_companion
    ),
  mcq_scores |>
    filter(amt %in% c(1, 3)) |>
    transmute(
      id, provider,
      amount = factor(if_else(amt == 1, "$30", "$80")),
      procedure = factor("MCQ"),
      log_k = log_k_companion
    )
)

companion_anova <- aov(
  log_k ~ provider * amount * procedure,
  data = companion_anova_data
)

summary(companion_anova)

companion_post_hoc <- emmeans(companion_anova, ~ procedure | amount) |>
  pairs(adjust = "tukey") |>
  as.data.frame()

kable(companion_post_hoc, digits = 3,
  caption = "Companion procedure contrasts within amount.")
```

### Published inferential results

| Effect | *F*(1, 1564) | *p* |
|---|---:|---:|
| Amount | 38.48 | < .001 |
| Procedure | 4.42 | .04 |
| Sample | 130.80 | < .001 |
| Procedure x Amount | 8.20 | .004 |

No other interactions were significant (all *p*s > .26). The Procedure x Amount interaction showed that the MCQ produced steeper discounting than Adj-Amt at $30 (*p* = .003), whereas the procedures did not differ at $80 (*p* = .95).

## 6. Recruitment-source differences and data quality

```{r distribution-summary}
participant_distributions <- full_join(
  adj_scores |>
    group_by(id, provider) |>
    summarise(mean_auc = mean(auc), .groups = "drop"),
  mcq_scores |>
    group_by(id, provider) |>
    summarise(total_delayed_choices = sum(delayed_choices), .groups = "drop"),
  by = c("id", "provider")
)

extreme_discounting <- participant_distributions |>
  group_by(provider) |>
  summarise(
    n = n(),
    `AuC < .125, n` = sum(mean_auc < .125),
    `AuC < .125, %` = 100 * mean(mean_auc < .125),
    `MCQ delayed choices <= 3, n` = sum(total_delayed_choices <= 3),
    `MCQ delayed choices <= 3, %` = 100 * mean(total_delayed_choices <= 3),
    `MCQ delayed choices = 0, n` = sum(total_delayed_choices == 0),
    `MCQ delayed choices = 0, %` = 100 * mean(total_delayed_choices == 0),
    .groups = "drop"
  )

kable(extreme_discounting, digits = 1,
  caption = "Extreme-discounting patterns by recruitment platform.")
```

No Prolific participant failed an attention check, whereas 32 of 275 MTurk recruits did. Even after those exclusions, the MTurk sample contained substantially more extremely steep discounters. These differences cannot be attributed causally to recruitment platform because the samples also differed in age, education, and ethnicity.

| Sample | Mean AuC < .125 | MCQ delayed choices <= 3 | No delayed MCQ choices |
|---|---:|---:|---:|
| Prolific | 1 (0.7%) | 7 (4.7%) | 0 (0.0%) |
| MTurk | 53 (21.8%) | 58 (23.9%) | 34 (14.0%) |

Counts and percentages in this table are recomputed from the analysis-ready dataset and reproduce the distributional statements in the article.

## 7. Interpretation and research guidance

| Research priority | Stronger practical fit | Reason |
|---|---|---|
| Brief online assessment | MCQ | Twenty-seven fixed items and established scoring |
| Rewards above $85 | Adj-Amt | Flexible delayed amounts |
| Individual discounting functions | Adj-Amt | Direct subjective-value estimates across delays |
| Probability, loss, or nonmonetary extensions | Adj-Amt or a separately validated questionnaire | The standard MCQ is specific to small delayed monetary gains |
| Comparing absolute `log k` values across studies | Neither without calibration | Procedure and amount can shift the level of `log k` |

### Conclusions

- Adj-Amt and MCQ show strong convergent validity across two online samples.
- The finding is robust to theoretical and atheoretical scoring approaches.
- The procedures are not automatically interchangeable at the level of absolute `log k`.
- Recruitment source and attention-check performance are part of the measurement context, not merely administrative details.
- Procedure choice should follow the scientific question, outcome domain, reward range, and required level of measurement detail.

### Scope and limitations

The evidence concerns delayed monetary gains in two United States online samples. It does not establish equivalence for delayed losses, probabilistic outcomes, combined delay-and-probability outcomes, nonmonetary rewards, or other populations. Correlations establish shared variance, not causal validity or score identity.

## References

Du, W., Green, L., & Myerson, J. (2002). Cross-cultural comparisons of discounting delayed and probabilistic rewards. *The Psychological Record, 52*, 479-492. <https://doi.org/10.1007/BF03395199>

Kirby, K. N., Petry, N. M., & Bickel, W. K. (1999). Heroin addicts have higher discount rates for delayed rewards than non-drug-using controls. *Journal of Experimental Psychology: General, 128*, 78-87. <https://doi.org/10.1037/0096-3445.128.1.78>

Myerson, J., Green, L., & Warusawitharana, M. (2001). Area under the curve as a measure of discounting. *Journal of the Experimental Analysis of Behavior, 76*, 235-243. <https://doi.org/10.1901/jeab.2001.76-235>

Wan, H., Myerson, J., & Green, L. (2023). Individual differences in degree of discounting: Do different procedures and measures assess the same construct? *Behavioural Processes, 208*, 104864. <https://doi.org/10.1016/j.beproc.2023.104864>
